---
title: "Ctrl+Alt+Analyze"
author: 
  - Malika
  - Kunal
  - ZuxiLu
format: 
  revealjs:
    theme: white 
    toc: false
    number-sections: true
    slide-number: true
    fig-cap-location: bottom
    transition: fade 
editor: visual
---

## Ctrl+Alt+Analyze

*By: Malika, Kunal, ZuxiLu*\
**ETC5513 – Collaborative & Reproducible Practices**

-   Rebooting learning through daily habits\
-   **Ctrl** your study time, **Alt** your distractions, **Del** your doubts.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(here)
library(corrplot)
library(kableExtra)
```

```{r}
#| label: load-data
#| echo: false
#| message: false
#| warning: false

data <- read_csv(here("data/student_habits_performance.csv"))
```

```{r}
#| label: clean-data
#| echo: false
#| message: false
#| warning: false

data_clean <- data |>
  rename(
    StudentID = student_id,
    Age = age,
    Gender = gender,
    StudyHours = study_hours_per_day,
    SocialMediaHours = social_media_hours,
    NetflixHours = netflix_hours,
    PartTimeJob = part_time_job,
    AttendanceRate = attendance_percentage,
    SleepHours = sleep_hours,
    DietQuality = diet_quality,
    ExerciseFrequency = exercise_frequency,
    ParentalEducation = parental_education_level,
    InternetQuality = internet_quality,
    MentalHealth = mental_health_rating,
    Extracurriculars = extracurricular_participation,
    ExamScore = exam_score
  )
```

## Problem Introduction

-   Academic performance is affected by daily habits like:

    -   Study hours per day
    -   Classroom attendance
    -   Sleep hours

-   On the other hand, spending too much time on:

    -   Social media
    -   Streaming platforms like Netflix

-   **Research Objective:**\
    Quantify the relationship between student habits and academic performance using correlation analysis.

-   **Goal:** Identify which habits support academic success and which ones hinder it.

## Dataset Description

-   100 students \| 16 variables \| Lifestyle + Exam performance

-   Main variables:

    -   Study hours per day
    -   Class attendance (%)
    -   Sleep hours
    -   Social media hours
    -   Netflix hours
    -   Exam score

## Grouped into:

-   **Good habits:** `StudyHours`, `AttendanceRate`, `SleepHours`

-   **Bad habits:** `SocialMediaHours`, `NetflixHours`

-   Target variable: `ExamScore`

-   Dataset was cleaned and renamed for clarity.

## Methods

-   **Correlational analysis** used to test how student lifestyle habits relate to academic performance

-   **Pearson correlations** calculated using `cor()` in R\

-   Visualized using **bar plots** with `ggplot2()`:

    -   **Blue** = Helps (positive correlation)\
    -   **Red** = Hurts (negative correlation)\
    -   Bars labeled with correlation values

## Table 1: Good Habits Correlation Table

```{r}
#| label: good-habits-score-correlation
#| tbl-cap: "Correlation of Study, Attendance, Sleep, and ExamScore"
#| echo: false
#| warning: false



# Select variables
vars <- c("StudyHours", "AttendanceRate", "SleepHours", "ExamScore")

# Compute correlation matrix
corr_matrix <- data_clean[, vars] |> cor(use = "complete.obs") |> round(2)

# Extract just correlation with ExamScore
corr_with_score <- corr_matrix[, "ExamScore", drop = FALSE]

# Remove column name to make it cleaner
colnames(corr_with_score) <- "Correlation with ExamScore"

# Display
kable(corr_with_score, caption = "Table 1: Correlation of Study, Attendance, Sleep, and ExamScore")
```

## Table 2: Bad Habits Correlation Table

```{r}
#| echo: false
#| results: asis
#| warning: false

bad_cor_matrix <- data_clean |>
  select(SocialMediaHours, NetflixHours, ExamScore) |>
  cor(use = "complete.obs") |>
  round(2)

kable(bad_cor_matrix, digits = 2) |>
  kable_styling(
    font_size = 24,         # Bigger font
    full_width = FALSE,     # Keeps it compact
    position = "center"     # Centers the table
  ) |>
  scroll_box(width = "90%", height = "auto")  # Allows wider layout
```

## Results

-   **StudyHours** has **strongest positive correlation** with exam scores (r = 0.83)
-   **AttendanceRate** and **SleepHours** show **weak positive correlations**
-   **SocialMediaHours** and **NetflixHours** each show **weak negative correlations**

## Results Continued

Bar plots below show how each habit helps or hurts academic performance:

```{r}
#| label: fig-bar-correlation
#| echo: false
#| warning: false
#| message: false

cor_df <- tibble(
  Habit = c("Study", "Attendance", "Sleep", "Social", "Netflix"),
  Correlation = c(0.83, 0.09, 0.12, -0.17, -0.17)
) |>
  mutate(
    Impact = ifelse(Correlation >= 0, "Helps", "Hurts"),
    LabelPos = ifelse(Correlation >= 0, Correlation - 0.05, Correlation + 0.05)
  )

ggplot(cor_df, aes(x = Habit, y = Correlation, fill = Impact)) +
  geom_col() +
  geom_text(
    aes(label = Correlation, y = LabelPos),
    color = "white",
    size = 5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("Helps" = "dodgerblue", "Hurts" = "firebrick")) +
  coord_cartesian(ylim = c(-0.25, 1)) +
  labs(
    x = "Student Habit",
    y = "Correlation with Exam Score (r)",
    fill = "Impact",
    title = "How Habits Correlate with Exam Performance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 16)
  )
```

## Conclusions & Recommendations

-   **StudyHours** is the most important factor (r = 0.83)
-   **Attendance** and **Sleep** have minor benefits
-   **Social media** and **Netflix** show mild negative impact

👉 Students should prioritize studying and manage distractions smartly.

👉 Data-driven habits can lead to better academic outcomes.
