---
title: "Ctrl+Alt+Analyze"
author: "Malika; Kunal; ZuxiLu"
format: revealjs
editor: visual
---

## Title slide

------------------------------------------------------------------------

## Problem introduction

-   Academic performance is affected by daily habits like:

    -   Study hours per day, Classroom attendance, Sleep hours

-   On the other hand, spending too much time on:

    -   Social media, Streaming platforms like Netflix\
        may reduce focus and study time.

-   **Research Objective:**\
    Quantify the relationship between student habits and academic performance using correlation analysis.

-   We aim to discover which habits support academic success and which habits hinder it.

------------------------------------------------------------------------

## Dataset description

-   The dataset includes **100 students** and **16 variables** covering lifestyle habits and exam performance.

-   Main variables include:

    -   Study hours per day\

    -   Class attendance (%)\

    -   Sleep hours\

    -   Social media hours\

    -   Netflix hours

    -   Exam score

------------------------------------------------------------------------

-   Main variables are grouped into:

    -   **Good habits:**\
        `StudyHours`, `AttendanceRate`, `SleepHours`
    -   **Bad habits:**\
        `SocialMediaHours`, `NetflixHours`

-   Target variable:\
    `ExamScore` (numeric score for student performance)

-   Data was cleaned and renamed for clarity before analysis.

------------------------------------------------------------------------

## Methods

### Analysis method

**Correlational analysis** used to test how student lifestyle habits relate to academic performance

### Technique

-   **Pearson correlations** drawn using `cor()` in R
-   Created **bubble-style** plots using `corrplot()` to visualize magnitude and direction of relationships [Blue]{style="color:blue"} = positive\
    [Red]{style="color:red"} = negative\
    Circle size = strength

```{r}
#| label: reading the data set for the first time
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(knitr)

data <- read_csv(here("data/student_habits_performance.csv"))

```

```{r}
#| label: cleaning the dataset a bit for further processing
#| echo: false
#| warning: false
#| message: false

data_clean <- data |>
  rename(
    StudentID = student_id,
    Age = age,
    Gender = gender,
    StudyHours = study_hours_per_day,
    SocialMediaHours = social_media_hours,
    NetflixHours = netflix_hours,
    PartTimeJob = part_time_job,
    AttendanceRate = attendance_percentage,
    SleepHours = sleep_hours,
    DietQuality = diet_quality,
    ExerciseFrequency = exercise_frequency,
    ParentalEducation = parental_education_level,
    InternetQuality = internet_quality,
    MentalHealth = mental_health_rating,
    Extracurriculars = extracurricular_participation,
    ExamScore = exam_score
  )

```

```{r}
#| echo: false
#| results: asis
#| layout-ncol: 2
#| label: side-by-side-correlation-tables

# Compute good habits correlation matrix
good_cor_matrix <- data_clean |>
  select(StudyHours, AttendanceRate, SleepHours, ExamScore) |>
  cor(use = "complete.obs") |>
  round(2)

# Compute bad habits correlation matrix
bad_cor_matrix <- data_clean |>
  select(SocialMediaHours, NetflixHours, ExamScore) |>
  cor(use = "complete.obs") |>
  round(2)

# Output side-by-side tables
cat("### Table 1: Good Habits and Exam Score Correlation\n\n")
print(kable(good_cor_matrix, digits = 2))

cat("\n\n### Table 2: Bad Habits and Exam Score Correlation\n\n")
print(kable(bad_cor_matrix, digits = 2))

```

------------------------------------------------------------------------

## Results

::: columns
-   **Study hours** shows a **strong positive correlation** with exam scores
-   **Attendance rate** and **Sleep hours** show only **weak positive correlation** with exam scores

```{r}
#| label: bubble-corrplot
#| fig-cap: "**Figure 1**: Bubble-style correlation matrix of student habits and exam score"

library(corrplot)

bubble_corr <- data_clean |>
  dplyr::select(StudyHours, AttendanceRate, SleepHours, ExamScore) |>
  cor(use = "complete.obs")

corrplot(
  bubble_corr,
  method = "circle",
  type = "lower",
  order = "hclust",
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("#d73027", "white", "#1E90FF"))(200),
  addCoef.col = "black",
  number.cex = 0.9,
  diag = FALSE
)

```
:::

::: columns
-   **Social media hours** and **Netflix hours** show only **weak negative correlation** with exam scores

```{r}
#| label: bubble-corrplot-bad
#| fig-cap: "**Figure 2**: Bubble-style correlation matrix of bad habits and exam score"

bad_bubble_corr <- data_clean |>
  dplyr::select(SocialMediaHours, NetflixHours, ExamScore) |>
  cor(use = "complete.obs") |>
  round(2)

colnames(bad_bubble_corr) <- c("Social", "Netflix", "Score")
rownames(bad_bubble_corr) <- c("Social", "Netflix", "Score")

par(mar = c(2, 2, 5, 2))  # Avoid clipping

corrplot(
  bad_bubble_corr,
  method = "circle",
  type = "lower",
  order = "original",
  tl.col = "black",
  tl.srt = 45,
  tl.cex = 0.9,
  cl.cex = 0.8,
  col = colorRampPalette(c("#d73027", "white", "#1E90FF"))(200),
  addCoef.col = "black",
  number.cex = 0.8,
  diag = FALSE
)
```
:::

------------------------------------------------------------------------

## Conclusions & recommendations
