---
title: "Assignment 3"
format: html
editor: visual
---

# Investigating the relationship between student's academic performance and lifestyle habits

## Executive summary

## Introduction

## Methodology

Loading required libraries

```{r}
#| label: Loading the required libraries 

library("here")
library("tidyverse")
library("ggplot2")
library("ggcorrplot")
```

Reading in the data

```{r}
#| label: reading the data set for the first time 

data <- read_csv(here("data", "student_habits_performance.csv"))

```

### Investigating 'good' habits

**Methodology**

We conducted a correlation analysis using numeric variables from our dataset to investigate the relationship between 'good' student lifestyle habits and academic performance. Specifically, we focused on three indicators of good lifestyle habits: `study_hours_per_day` , `attendance_percentage` , and `sleep_hours` , alongside the outcome variable `exam_score`.

First, we filtered the dataset to retain these four variables that are related to our analysis. Then, we computed a Pearson correlation matrix and accounted for any missing values. We then visualised the correlation between `exam_score` and the other predictive variables in order to aid our analysis of which study habits are most strongly correlated with academic success.

```{r}

# Selecting relevant variables 

good_habits <- data |>
  select(study_hours_per_day, attendance_percentage, sleep_hours, exam_score)

# Creating a correlation matrix for these variables 
good_cor_matrix <- cor(good_habits , use = "complete.obs")

# Visualising correlation to good habits 
ggcorrplot(good_cor_matrix, method = "circle", type = "lower",
           lab = TRUE, lab_size = 4, title = "Figure 1: Correlation Matrix of 'Good' Student Habits and Exam Score")
```

### Investigating 'bad' habits

## Results

## Discussion

## References
